Does freshwater connectivity influence phosphorus retention in lakes?

### 2019-01-22 manuscript/manuscript.Rmd
fill out refs, enlargen last figs

In general, point source inputs to lakes are associated with increased dissolved P loads \citep{kronvang1992, russell1998}, whereas non-point source inputs to lakes, such as those in lake watersheds with high agricultural land use cover, typically have higher particulate P loads \citep{carpenter1998, sharpley1994}. Exceptions to these generalizations have been observed in areas where increased dissolved P loading occurs not as a result of point source nutrient inputs but rather as a result of non-point source runoff due to P saturated soils or legacy P release \citep{bennet2001, powers2015}. Despite the fact that particulate loads are usually related to non-point source inputs, the fraction of non-point source inputs that are in particulate form is highly variable and may depend on intra-annual flow variations as well as watershed erosion characteristics \citep{jarvie2011}. Furthermore, quantifying non-point source inputs at broad spatial scales is not commonly done due to logistical and sampling constraints \citep{guy1994}. As a result, studies often infer the relative amounts of particulate and dissolved loading from other available proxy data such as land-use cover (Ellison and Brett 2006, Djodjic and Markensten 2018).

Apart from land-use cover, another potential proxy for particulate load fraction entering lakes is the type and amount of freshwater connections to a lake, which we argue is easier to measure than other proxies, and could help to improve estimates of lake P retention, especially in lakes for which we lack P loading data (Figure 1). Although freshwater connectivity may be easy to measure from a logistical standpoint, there are still many ways to measure connectivity that likely represent different mechanisms of water and material flow (Figure 2). We broadly define and study two types of freshwater connectivity that correspond to either stream-based metrics or lake-stream-based metrics. First, lake-stream-based metrics measure the connections between a lake, other upstream lakes, and streams in their watershed. This type of connectivity can be quantified by measuring the closest distance to an upstream stream-connected lake (Figure 2A), or by measuring the total upstream lake area (Figure 2B). Second, stream-based metrics measure the connections between inflowing streams and their surrounding land whereby increasing stream connections lead to a greater abundance of land-water interfaces and greater transport of material from hillslope and riparian areas adjacent to watershed stream networks (Figure 2C-E).

@@

In addition to variation among connectivity metrics and connectivity metric types, it is also not always clear which spatial extent is appropriate for determining functional differences in connectivity among lakes (Soranno et al. 2015). Such information is needed to inform the design of regulatory frameworks balancing controls on cumulative nutrient transport along stream networks and controls on localized nutrient transport (Alexander et al. 2008; Carpenter et al. 1998; Withers and Jarvie 2008). To test the importance of different spatial extents, we examined connectivity metrics measured for both the lake subwatershed extent and the lake watershed extent (Figure 3). Here, the lake subwatershed extent includes the elements of the immediate watershed in the direct drainage of a lake whereas the lake watershed extent includes all of the elements in the entirety of the lake-stream network up to and including headwater streams (Figure 3).

@@

We propose that measures of freshwater connectivity are related to P retention in the following ways. First, some connectivity metrics reflect the proximity of terrestrial watershed areas to the stream network (Covino 2017). For example, lakes with a high watershed stream density should have increased particulate matter loading from terrestrial hillslope and riparian areas adjacent to watershed stream networks because particulate matter export is limited by overland distance (Gomi et al. 2002). Second, connectivity metrics can reflect the configuration of lakes within lake-stream networks. For example, lakes with upstream lakes in close proximity may receive P loads that have previously undergone in-lake processing whereby labile fractions have already been trapped in upstream lakes (Cardille et al. 2007). In contrast, lakes with more distant upstream lakes are more likely to receive the more labile fractions from terrestrial runoff that serve to increase P retention as opposed to receiving the more recalcitrant fractions that are resistant to biological uptake and are thus not retained. Although some connectivity metrics have an intuitive relation to P retention, it is not clear which specific measures of freshwater connectivity are important for transport of particulate matter. Therefore, our study is designed to examine and compare which measures of connectivity are more related to lake P retention.

@@

\begin{figure}
  \begin{center}
    \includegraphics[width=0.5\textwidth,keepaspectratio]{../figures/08_maps-1}
  \end{center}
  \caption{Locations of lakes of with different connectivity metric values at the (A) lake subwatershed and (B) lake watershed extent. Low valued lakes are represented by lighter (green) symbols, and high values lakes are represented by darker (purple) symbols. WS is the lake watershed extent. SWS is the lake subwatershed extent. Connectivity metrics are defined in Figure 2.}\label{fig5}
\end{figure}

@@

\begin{figure}
  \begin{center}
    \includegraphics[width=0.5\textwidth,keepaspectratio]{../figures/06_k-3}
  \end{center}
  \caption{Distribution of the k parameter from Eq. 1 in lakes of differing connectivity, depth, or baseflow at the (A) lake subwatershed and (B) lake watershed extents. Lighter (green) lines indicate lakes with lower connectivity metrics values while darker (purple) lines indicate lakes with higher connectivity metrics values. For lake connection, lighter colored lines indicate lakes without upstream lakes. Connectivity metrics are defined in Figure 2. Labels associated with models where differences in k translated to significant differences in P retention are bolded and starred.}\label{fig7}
\end{figure}

>>>>

In general, point source inputs to lakes are associated with increased dissolved P loads \citep{kronvang1992, russell1998}, whereas non-point source inputs to lakes, such as those in lake watersheds with high agricultural land use cover, typically have higher particulate P loads \citep{carpenter1998, sharpley1994}. Exceptions to these generalizations have been observed in areas where increased dissolved P loading occurs not as a result of point source nutrient inputs but rather as a result of non-point source runoff due to P saturated soils or legacy P release \citep{bennet2001, powers2015}. Despite the fact that particulate loads are usually related to non-point source inputs, the fraction of non-point source inputs that are in particulate form is highly variable and may depend on intra-annual flow variations as well as watershed erosion characteristics \citep{jarvie2011}. Furthermore, quantifying non-point source inputs at broad spatial scales is not commonly done due to logistical and sampling constraints \citep{guy1994}. As a result, studies often infer the relative amounts of particulate and dissolved loading from other available proxy data such as land-use cover \citep{ellisonParticulatePhosphorusBioavailability2006, djodjicSingleFieldsRiver2018}.

Apart from land-use cover, another potential proxy for particulate load fraction entering lakes is the type and amount of freshwater connections to a lake, which we argue is easier to measure than other proxies, and could help to improve estimates of lake P retention, especially in lakes for which we lack P loading data (Figure \ref{fig1}). Although freshwater connectivity may be easy to measure from a logistical standpoint, there are still many ways to measure connectivity that likely represent different mechanisms of water and material flow (Figure \ref{fig2}). We broadly define and study two types of freshwater connectivity that correspond to either stream-based metrics or lake-stream-based metrics. First, lake-stream-based metrics measure the connections between a lake, other upstream lakes, and streams in their watershed. This type of connectivity can be quantified by measuring the closest distance to an upstream stream-connected lake (Figure \ref{fig2}A), or by measuring the total upstream lake area (Figure \ref{fig2}B). Second, stream-based metrics measure the connections between inflowing streams and their surrounding land whereby increasing stream connections lead to a greater abundance of land-water interfaces and greater transport of material from hillslope and riparian areas adjacent to watershed stream networks (Figure \ref{fig2}C-E).

@@

In addition to variation among connectivity metrics and connectivity metric types, it is also not always clear which spatial extent is appropriate for determining functional differences in connectivity among lakes \citep{soranno2015}. Such information is needed to inform the design of regulatory frameworks balancing controls on cumulative nutrient transport along stream networks and controls on localized nutrient transport \citep{alexander2008, carpenter1998, withersDeliveryCyclingPhosphorus2008}. To test the importance of different spatial extents, we examined connectivity metrics measured for both the lake subwatershed extent and the lake watershed extent (Figure \ref{fig3}). Here, the lake subwatershed extent includes the elements of the immediate watershed in the direct drainage of a lake whereas the lake watershed extent includes all of the elements in the entirety of the lake-stream network up to and including headwater streams (Figure \ref{fig3}).

@@

We propose that measures of freshwater connectivity are related to P retention in the following ways. First, some connectivity metrics reflect the proximity of terrestrial watershed areas to the stream network \citep{covino2017}. For example, lakes with a high watershed stream density should have increased particulate matter loading from terrestrial hillslope and riparian areas adjacent to watershed stream networks because particulate matter export is limited by overland distance \citep{gomi2002}. Second, connectivity metrics can reflect the configuration of lakes within lake-stream networks. For example, lakes with upstream lakes in close proximity may receive P loads that have previously undergone in-lake processing whereby labile fractions have already been trapped in upstream lakes (Cardille et al. 2007). In contrast, lakes with more distant upstream lakes are more likely to receive the more labile fractions from terrestrial runoff that serve to increase P retention as opposed to receiving the more recalcitrant fractions that are resistant to biological uptake and are thus not retained. Although some connectivity metrics have an intuitive relation to P retention, it is not clear which specific measures of freshwater connectivity are important for transport of particulate matter. Therefore, our study is designed to examine and compare which measures of connectivity are more related to lake P retention.

@@

\begin{figure*}
  \begin{center}
    \includegraphics[width=0.6\textwidth,keepaspectratio]{../figures/08_maps-1}
  \end{center}
  \caption{Locations of lakes of with different connectivity metric values at the (A) lake subwatershed and (B) lake watershed extent. Low valued lakes are represented by lighter (green) symbols, and high values lakes are represented by darker (purple) symbols. WS is the lake watershed extent. SWS is the lake subwatershed extent. Connectivity metrics are defined in Figure 2.}\label{fig5}
\end{figure*}

@@

\begin{figure*}
  \begin{center}
    \includegraphics[width=0.9\textwidth,keepaspectratio]{../figures/06_k-3}
  \end{center}
  \caption{Distribution of the k parameter from Eq. 1 in lakes of differing connectivity, depth, or baseflow at the (A) lake subwatershed and (B) lake watershed extents. Lighter (green) lines indicate lakes with lower connectivity metrics values while darker (purple) lines indicate lakes with higher connectivity metrics values. For lake connection, lighter colored lines indicate lakes without upstream lakes. Connectivity metrics are defined in Figure 2. Labels associated with models where differences in k translated to significant differences in P retention are bolded and starred.}\label{fig7}
\end{figure*}
